### 2023-08-31 sections/2-methods.tex
getting there

In this manuscript, we are primarily concerned with $k$-nearest neighbors in a finite-dimensional 
vector space.

Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$, we define a \emph{point} or \emph{datum} $x_i \in \textbf{X}$ as a singular observation (e.g., the genome of 
an organism, the vector representation of a single image, or any entity on which we can define a \emph{distance function}).

We define a \emph{distance function} $f : (\textbf{X}, \textbf{X}) \mapsto \mathbb{R}^+ \cup \{0\}$ as a function which, 
given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number. We require that the distance function 
be symmetric ($f(x, y) = f(y, x)$ for all $x, y \in \textbf{X}$) and that the distance between two points $x$ and $y$ be zero if and only if $x = y$. 
When, in addition to these constraints, the distance function obeys
the triangle inequality, it is a \emph{distance metric}, and in such cases we can guarantee that our search algorithms have perfect recall. 
Choice of distance function varies by type of data. For example, with electromagnetic spectra, we use both 
Euclidean (L2) and Cosine distances. With biological or string data, Hamming and Levenshtein distances are more appropriate.

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} at some point in the dataset, 
which we define as: 
\begin{equation} \frac{\text{log}(\frac{|B_X(q, r_1)|}{|B_X(q, r_2)|})}{\text{log}(\frac{r_1}{r_2}) } \label{1} \end{equation}
where $B_X(q, r)$ is the set of points contained in a ball of radius $r$ 
centered at a point $q$ in the dataset $\textbf{X}$; in this example, we compute fractal dimension for some radius $r_1$ and a smaller radius $r_2 = \frac{r_1}{2}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset. To illustrate the difference,
consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575. Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional 
manifold within the 8,575-dimensional embedding space. The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold at a given point, for some length scale.
The notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold is known as the \emph{manifold hypothesis}.
[TODO: add example for GreenGenes and add lfd plots from CHESS]

@@

With these concepts defined, we can now pose the $k$- and $\rho$ nearest neighbors search problems.
Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search aims to find 
the set $\{x \in \textbf{X}: f(q, x) \leq f(q, x_k) \land x_k \text{ is the $k$th nearest neighbor of $q$ in $\textbf{X}$}\}$; that is,
for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$.
We also have the $\rho$-nearest neighbors search problem, which aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$; that is, 
find all points in $\textbf{X}$ that at most a distance $\rho$ from $q$.

@@

We define \emph{singletons} as clusters which either contain a single point (i.e., have cardinality 1) or which contain 
many instances of the same point (i.e., have cardinality greater than 1 but contain only one unique point). Singletons clearly 
have zero radius, and so $\delta_{max} = \delta_{min} = \delta$ for these clusters. Hence, we sometimes overload the above 
notation to refer to distance from a query to a point. Similarly, we have that $\delta_{max} = \delta = \delta_{min}$ for 
individual points.

@@

For a cluster $C$ with $|C|$ points, we begin by computing a 
random sample of $\sqrt{|C|}$ points, and computing pairwise distances 
between all points in the sample. Based on those pairwise distances, we compute the \emph{geometric median} of this sample; 
that is, the point which minimizes the sum of distances to all other points in the sample. This geometric median 
is $C$'s \emph{center}. The \emph{radius} of $C$ is the maximum distance from any point in $C$ to its center.
The point $l$ which is furthest from the center is designated the \emph{left pole}, and the point $r$ which is furthest
from $l$ is designated the \emph{right pole}. We then partition the cluster into a \emph{left child} and \emph{right child}, where the 
left child contains all points in the cluster which are closer to $l$ than to $r$, and the right child contains all 
points in the cluster which are closer to $r$ than to $l$. Without loss of generality, we assign to the left child 
those points which are equidistant from $l$ and $r$. Starting from a root-cluster containing the entire dataset, we 
repeat this procedure until each leaf contains only one datum or until some other user-specified stopping criterion 
is met.

\begin{algorithm} % enter the algorithm environment
\caption{Partition} % give the algorithm a caption
\label{alg:partition} % and a label for \ref{} commands later in the document
\begin{algorithmic}[1] % enter the algorithmic environment
    \REQUIRE $cluster$
    \STATE $m \leftarrow \lfloor \sqrt{|cluster.points|} \rfloor$
    \STATE $seeds \leftarrow m$ random points from $cluster.points$
    \STATE $c \leftarrow$ geometric median of $seeds$
    \STATE $l \leftarrow \argmax d(c,x) \ \forall \ x \in cluster.points$
    \STATE $r \leftarrow \argmax d(l,x) \ \forall \ x \in cluster.points$
    \STATE $left \leftarrow \{x | x \in cluster.points \land d(l,x) \le d(r,x)\}$
    \STATE $right \leftarrow \{x | x \in cluster.points \land d(r,x) < d(l,x)\}$
    \IF{$|left| > 1$}
        \STATE Partition($left$)
    \ENDIF
    \IF{$|right| > 1$}
        \STATE Partition($right$)
    \ENDIF
\end{algorithmic}
\end{algorithm}

@@

CLAM-CAKES also reorders the dataset to improve the tree-building process 
used in CHESS. In CHESS, each cluster stored a vector of indices of its points. 
We use these indices during search to retrieve a cluster's points from the dataset. 
Though this approach allowed us to retrieve the points in constant 
time, its memory cost was prohibitively high; since each cluster stores indices of all of its
points, for a dataset with cardinality $n$, we store a total of $n$ indices at each depth in the tree.
With $\log{}n$ levels in the tree, this approach had $\mathcal{O}(n\log{}n)$ memory cost.

A later improvement reduced the memory cost to $\mathcal{O}(n)$ by storing only the indices of the points in the leaves.
This approach, however, introduced an $\mathcal{O}(n)$ time cost every time we needed to find the indices for 
a non-leaf cluster, since this required a traversal of the subtree rooted at that cluster.

In this work, we introduce a new approach, wherein after building the cluster tree, we reorder the dataset 
so that points are stored in a depth-first order based on the tree. Then, within each cluster, we need only 
store the cluster's cardinality, and an \emph{offset} to access the cluster's points from the dataset.
For left leaf clusters, the offset is the index of the first point in the cluster. Right leaves have an offset 
equal to their sibling's offset plus the their sibling's cardinality. Parents have the same offset as their left child.
With $\mathcal{O}(n)$ memory cost and $\mathcal{O}(1)$ time cost for retrieving a points during search, 
dataset depth-first reordering offers significant improvements over the previous approaches.

\subsubsection {Guaranteed Decrease in Cluster Radii}
\label{subsubsec:methods:guaranteed-decrease-in-cluster-radii}

TODO: This needs some doctoring.

We show that cluster radius is guaranteed to decrease after at most $d$ partitions.

Assume that we have a dataset which follows a $d$-dimensional distribution embedded in $D$-dimensional space.
We can describe this distribution choosing some set of $d$ mutually orthogonal axes.
Let $m$ denote the maximum distance between any pair of points along any axis from any choice of mutually orthogonal axes for our $d$-dimensional distribution.
In the worst case scenario, the distribution of our data is a $d$-sphere and $m$ is the maximum distance between two points along every axis for every choice of axes.
In this scenario, the cluster radius is $r = \frac{m}{2}$.

Given a cluster $C$, we choose the left and right poles, 
and the axis defined by those poles will be the axis for partitioning  $C$'s points into two child clusters.
For points in each child cluster, the maximum distance between any pair of points along the \emph{axis of partitioning} will have been reduced to at most $\frac{m}{2}$.
Though this bound holds for points on the axis of partitioning, it does not serve as a bound for the maximum distance between \emph{any} pair of points in  $C$.
To see this, consider the two dimensional scenario where the points in  $C$ are distributed in a circle of radius $r$. Clearly, the maximum possible distance between 
any two points in  $C$ is $2r$, and occurs when the points fall on opposite ends of one of its diameters. Now consider a horizontal diameter of the circle, 
and let $l$ denote the leftmost point on this diameter. Next, consider the orthogonal vertical diameter, and let $x$ denote the uppermost point on this diameter. When we 
partition along the horizontal diameter, $l$ and $x$ will be assigned to the left child (as mentioned in Section 2, we assign equidistant points to the left child). 
This reduces the maximum possible distance between any two points on the axis of partition --the horizontal diameter-- to $r$. However, by considering the 
right triangle formed by $l$, $x$, and the center of the circle, we can see that the distance between 
$l$ and $x$ is $\sqrt{2}{r}$.

If $d > 1$ then the child clusters will have a maximal chord $m_{child}$ such that $m_{child} \leq m$.
By partitioning the cluster in this way, we have taken one axis and reduced the maximum distance along that axis from $m$ to $\frac{m}{2}$.
However, as per the distribution of data as discussed earlier, there may be other axes along which the maximum pairwise distance is still $m$.
We, thus, recursively partition each child cluster.

With each recursive application of Partition, we consume one axis along which the maximum distance was $m$ and reduce the maximum distance along that axis to $\frac{m}{2}$ in the child clusters.
After $d$ recursive applications of partition, we will have exhausted the $d$ axes with the maximum distance of $m$.
After these $d$ partitions, the points in each child cluster will have a maximum pairwise distance of $\frac{m}{2}$ along any axis.
Thus the radii of those child clusters will be at most $\frac{m}{4}$.

\subsubsection {Complexity}
\label{subsubsec:methods:clustering:complexity}

Clustering: Exact partition with $\mathcal{O}(n^2)$ cost vs approximate partition using $\sqrt{n}$ seeds to achieve $\mathcal{O}(n)$ cost.
Building the exact tree costs $\mathcal{O}(n^2 \log n)$ vs approximate tree for $\mathcal{O}(n \log n)$.

\subsection {Index-Building}
\begin{itemize}
    \item sharded search
    \item index building 
    \item optimization of number of shards 
\end{itemize}

@@

[TODO: reorder so rho nn is first, then sieves, then Greedy Sieve]
In this section, we present four novel algorithms for $k$-nearest neighbors search: Greedy Search, $k$-NN by Repeated $\rho$-NN, Sieve Search I, and Sieve Search II. 
We also have an implementation of linear search. We use a preliminary auto-tuning function to predict the variant which will perform 
best for a given query, dataset, and value of $k$. We then proceed with search using that variant.

[TODO: tighten up or omit those opening paragraphs which roughly explain the algorithm]

[TODO: call queues of points H]
\subsubsection{Greedy Search}
\label{subsubsec:methods:knn-search:greedy-search}
Greedy Search repeatedly partitions the cluster with the smallest $\delta_{min}$ into its children 
until the cluster with the smallest $\delta_{min}$ is a leaf. We then remove it and add its points to a fixed size priority queue of points. 
Search terminates when the furthest hit in the queue of points has $\delta$ greater than the $\delta_{min}$ of the closest remaining cluster.

Formally, let $Q$ be a priority queue of clusters sorted by non-increasing $\delta_{min}$. Initialize $Q$ with the root cluster.
Let $P$ be a second priority queue of points sorted by non-decreasing distance from the query. This queue starts empty.

[TODO: replace with pop and push language and start with initial pop]
While $P$ has fewer than $k$ clusters or $\delta$ of the top priority element in $P$ is greater 
than $\delta_{min}$ of the top priority element in $Q$, do the following:
\begin{enumerate}
\item While the top priority element $q$ in $Q$ is not a leaf, replace $q$ with its children.
\item Remove $q$ from $Q$ and add all of its points to $P$. 
\item Remove points from $P$ until $P$ has $k$ points. 
\end{enumerate}
After this process, the points left in $P$ are the $k$ nearest neighbors of $q$.

[TODO: fix to use priority queue language]

\begin{algorithm} 
\caption{GreedySearch(\emph{tree, query, k})} 
\label{alg:greedy_search} 
\begin{algorithmic}[3]
    \STATE $Q \leftarrow$ priority queue
    \STATE $Q.push(tree.root)$
    \STATE $P \leftarrow$ priority queue
    \WHILE{$|P| < k \lor P.max.\delta > Q.min.\delta_{min}$}
        \WHILE{$\neg Q.min.isLeaf$}
            \STATE $[l, r] \leftarrow Q.extractMin().children$
            \STATE $Q.push([l, r])$
        \ENDWHILE
        \STATE $leaf \leftarrow Q.extractMin()$
        \STATE $P.push(leaf)$
        \WHILE{$|P| > k$}
            \STATE $P.extractMax()$
        \ENDWHILE
    \ENDWHILE
    \STATE Return $P$
\end{algorithmic}
\end{algorithm}

\subsubsection{$k$-NN by Repeated $\rho$-NN}
\label{subsubsec:methods:knn-search:repeated-rnn}
In this algorithm, we perform
$\rho$-nearest neighbors search at [TODO: add radius], repeatedly increasing the search radius until $k$ neighbors
have been found.

[TODO: change P to H]
Let $P$ be the set of nearest neighbors found thus far.
Search starts with a radius $m$ equal to the radius of the cluster tree divided by
the cardinality of the dataset. We perform $\rho$-NN search with radius $m$. 
If no points are within a distance $m$ of the query, we increase the radius by a factor of 
2 and perform $\rho$-NN search again, repeating until at least one point is found, i.e., 
until $|P| > 0$.

[TODO: Add number next to all equations and fix factor so that it is correct (k should be raised too)]
Once $|P| > 0$, we continue to perform $\rho$-NN search, but instead of 
increasing the radius by a factor of 2 on each iteration, we increase it by a factor determined 
by the local fractal dimension of the clusters containing the points found so far. In particular, 
we increase the radius by a factor of 
$$\text{min}\left(2, \frac{k}{|P|^{\frac{1}{\mu}}}\right)$$
where $\mu$ is the mean local fractal dimension (abbreviated lfd in algorithm below) of the clusters containing the points found so far.
Intuitively, the factor by which we increase the radius should be \emph{inversely} proportional to the number of points found so far. 
Additionally, when the local fractal dimension near the query and found neighbors is low, this suggests that the data 
are relatively concentrated in that region; thus, the factor of radius increase should be \emph{directly} proportional to the 
local fractal dimension. Since the local fractal dimension is memoized at clustering time, this poses no additional cost at search time.
Once $|P| >= k$, we sort the points in $P$ by $\delta$ and return the $k$ closest points.

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated$\rho$-NN(\emph{tree, query, k})} % give the algorithm a caption
    \label{alg:knn-by-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[4] % enter the algorithmic environment
        \STATE $P \leftarrow$ $\emptyset$
        \STATE $m \leftarrow$ $\frac{tree.radius}{tree.cardinality}$
        \WHILE {$|P| = 0$}
            \STATE $P.push(\rho$-NN$(tree, query, m)$)
            \STATE $m \leftarrow 2m$
        \ENDWHILE
        \WHILE {$|P| < k$}
            \STATE $clusters \leftarrow \{ C: \exists p \in P \land p \in C \}$
            \STATE $\mu \leftarrow \frac{1}{|clusters|} \sum_{C \in clusters} C.lfd$
            \STATE $m \leftarrow \text{min}\left(2, \frac{k}{|P|^{\frac{1}{\mu}}}\right)$
            \STATE $P.push(\rho$-NN$(tree, query, m)$)
        \ENDWHILE
        \STATE $P.sort()$
        \STATE Return $P[0..k]$
    \end{algorithmic}
    \end{algorithm}

[TODO: change sieve II to sieev with separate centers]
\subsubsection{Sieve Search I}
\label{subsubsec:methods:knn-search:sieve}
[TODO: make writeup here tighter and more consistent with what's in the algorithm; talk 
about enum variants here]
With Sieve Search I, we repeatedly calculate a threshold distance $\tau$ such that no cluster with a $\delta_{min}$ greater than $\tau$ 
can contain one of the $k$ nearest neighbors. Search terminates when we have a threshold containing exactly $k$ points.

We start by letting $Q$ be a vector initialized with the root cluster. This vector will maintain the 
clusters and points which are still in contention for being one of the nearest neighbors.
While $Q$ contains at least one cluster (i.e., $Q$ is not a vector of only points), we repeat the process described in 
the following paragraphs.

@@

We complete this process until $Q$ contains only points. At this point, we 
just use [INSERT ref to quicksort partition in paper] to find the $k$th 
nearest neighbor. Since this algorithm will also elements on the correct side of 
the index of the $k$th nearest neighbor, the first $k$ elements in $Q$ a are the $k$-nearest neighbors.

>>>>

In this manuscript, we are primarily concerned with $k$-nearest neighbors in a finite-dimensional 
vector space. Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$, we define a \emph{point} or \emph{datum} $x_i \in \textbf{X}$ as a singular observation (e.g., the genome of 
an organism, the vector representation of a single image, or any entity on which we can define a \emph{distance function}).

We define a \emph{distance function} $f : (\textbf{X}, \textbf{X}) \mapsto \mathbb{R}^+ \cup \{0\}$ as a function which, 
given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number. We require that the distance function 
be symmetric (i.e., $f(x, y) = f(y, x)$ for all $x, y \in \textbf{X}$) and that the distance between two points $x$ and $y$ be zero if and only if $x = y$. 
When, in addition to these constraints, the distance function obeys
the triangle inequality, it is a \emph{distance metric}, and in such cases we can guarantee that our search algorithms have perfect recall. 
Choice of distance function varies by type of data. For example, with electromagnetic spectra, we use both 
Euclidean (L2) and Cosine distances. With biological or string data, Hamming and Levenshtein distances are more appropriate.

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} at some point in the dataset, 
which we define as: 
\begin{equation} \frac{\text{log}(\frac{|B_X(q, r_1)|}{|B_X(q, r_2)|})}{\text{log}(\frac{r_1}{r_2}) } \label{1} \end{equation}
where $B_X(q, r)$ is the set of points contained in a ball of radius $r$ 
centered at a point $q$ in the dataset $\textbf{X}$; in this example, we compute fractal dimension for some radius $r_1$ and a smaller radius $r_2 = \frac{r_1}{2}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset. To illustrate the difference,
consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575. Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional 
manifold within the 8,575-dimensional embedding space. The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold at a given point, for some length scale.
The notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold is known as the \emph{manifold hypothesis}.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=2.5in]{images/lfd_plots/lfd_plot_apogee.png}
    \caption{Mean fractal dimension of APOGEE clusters as a function of
    depth when clustered based on L2 norm. Each plot line represents a distinct
    decile of fractal dimension. Beyond a depth of 56, no clusters are further
    divided due to the minimum-cluster-cardinality stopping criteria.}
    \label{fig:methods:lfd_apogee}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics[width=2.5in]{images/lfd_plots/lfd_plot_greengenes.png}
    \caption{Mean fractal dimension of GreenGenes clusters as a function of
    depth when clustered based on Hamming distance. Each plot line represents
    a distinct decile of fractal dimension.}
    \label{fig:methods:lfd_greengenes}
\end{figure}

Figure~\ref{fig:methods:lfd_apogee} shows the mean local fractal dimension of the APOGEE dataset at each level of hierarchical clustering, 
under the L2 (Euclidean) distance metric. Each plot line represents a decile of fractal dimension; note that, other than the most extreme 10\% of clusters,
virtually all clusters have a local fractal dimension of less than $2 \ll 8,575$. This suggests that APOGEE is a good candidate for use with CAKES.

Figure~\ref{fig:methods:lfd_greengenes} shows the mean local fractal dimension of the GreenGenes dataset at each level of hierarchical clustering,
under the Hamming distance metric. Each plot line represents a decile of fractal dimension; once again, other than the most extreme 10\% of clusters,
virtually all clusters have a local fractal dimension of less than $2 \ll 2,250$. This suggests that GreenGenes is a good candidate for use with CAKES.

@@

With these concepts defined, we can now pose the $k$- and $\rho$- nearest neighbors search problems.
Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search aims to find 
the set $S_q$ such that  $|S_q| = k$ and $\forall x \in X \setminus S_q$, $f(q, x) < \max\{f(y, q): y \in S_q \}$; that is,
for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$.
We also have the $\rho$-nearest neighbors search problem, which aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$; that is, 
find all points in $\textbf{X}$ that are at most a distance $\rho$ from $q$.

@@

We define \emph{singletons} as clusters which either contain a single point (i.e., have cardinality 1) or which contain 
many instances of the same point (i.e., have cardinality greater than 1 but contain only one unique point). Singletons clearly 
have zero radius, and so $\delta_{max} = \delta_{min} = \delta$ for these clusters. Hence, we sometimes overload the above 
notation to refer to distance from a query to a point; in these cases, we also have that $\delta_{max} = \delta = \delta_{min}$.

@@

\subsubsection {Cluster Partitioning}

For a cluster $C$ with $|C|$ points, we begin by computing a 
random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances 
between all points in the sample. Based on those pairwise distances, we compute the \emph{geometric median} of this sample; 
that is, the point which minimizes the sum of distances to all other points in the sample. This geometric median 
is $C$'s \emph{center}. The \emph{radius} of $C$ is the maximum distance from any point in $C$ to its center.
The point $l$ which is responsible for that radius (i.e., the furthest point from the center) is designated the \emph{left pole}, and the point $r$ which is furthest
from $l$ is designated the \emph{right pole}. We then partition the cluster into a \emph{left child} and \emph{right child}, where the 
left child contains all points in the cluster which are closer to $l$ than to $r$, and the right child contains all 
points in the cluster which are closer to $r$ than to $l$. Without loss of generality, we assign to the left child 
those points which are equidistant from $l$ and $r$. Starting from a root-cluster containing the entire dataset, we 
repeat this procedure until each leaf contains only one datum or until some other user-specified stopping criterion 
is met.

\begin{algorithm} % enter the algorithm environment
\caption{Cluster Partition} % give the algorithm a caption
\label{alg:partition} % and a label for \ref{} commands later in the document
\begin{algorithmic}[1] % enter the algorithmic environment
    \REQUIRE $cluster$
    \STATE $m \leftarrow \lfloor \sqrt{|cluster.points|} \rfloor$
    \STATE $seeds \leftarrow m$ random points from $cluster.points$
    \STATE $c \leftarrow$ geometric median of $seeds$
    \STATE $l \leftarrow \argmax d(c,x) \ \forall \ x \in cluster.points$
    \STATE $r \leftarrow \argmax d(l,x) \ \forall \ x \in cluster.points$
    \STATE $left \leftarrow \{x | x \in cluster.points \land d(l,x) \le d(r,x)\}$
    \STATE $right \leftarrow \{x | x \in cluster.points \land d(r,x) < d(l,x)\}$
    \IF{$|left| > 1$}
        \STATE Partition($left$)
    \ENDIF
    \IF{$|right| > 1$}
        \STATE Partition($right$)
    \ENDIF
\end{algorithmic}
\end{algorithm}

@@

CAKES also improves upon CHESS by reordering the dataset. In CHESS, each cluster stored a list of indices into the dataset, 
which we used during search to retrieve a cluster's points from the dataset. 
Though this approach allowed us to retrieve the points in constant 
time, its memory cost was prohibitively high; since each cluster stores indices for each of its
points, for a dataset with cardinality $n$, we store a total of $n$ indices at each depth in the tree.
With $\log{}n$ levels in the tree, this approach had $\mathcal{O}(n\log{}n)$ memory cost.

One may think to improve this memory cost to $\mathcal{O}(n)$ by storing only the indices of points at 
the leaf cluster level.
This approach, however, introduces an $\mathcal{O}(n)$ time cost whenever we need to find the indices for 
a non-leaf cluster, since it a traversal of the subtree rooted at that cluster.

In this work, we introduce a new approach, wherein after building the cluster tree, we reorder the dataset 
so that points are stored in a depth-first order. Then, within each cluster, we need only 
store the cluster's cardinality and an \emph{offset} to access the cluster's points from the dataset. Parent clusters 
have the same offset as their left child, and right child clusters have an offset equal to their sibling's offset
plus their siblings's cardinality. With no additional memory cost nor time cost for retrieving a points during search, 
dataset depth-first reordering offers significant improvements over the approach used in CHESS.

\subsubsection {Scaling Behavior of Cluster Radii}
\label{subsubsec:methods:guaranteed-decrease-in-cluster-radii}
While it may be tempting to assume that cluster radii decrease with each application of Cluster Partition (TODO: ADD SECTION), unfortunately, this assumption is incorrect. 
Fortunately, we \emph{can} make some guarantees about the scaling behavior of cluster radii; in particular, we can guarantee that cluster radii will stop increasing after at most 
$d$ partitions, where $d$ is the dimensionality of the dataset.  In the remainder of this subsection, we HJFHFJ.

We can describe this a $d$-dimensional distribution by choosing some set of $d$ mutually orthogonal axes.
Let $2R$ be the maximum pairwise distance among the instances in the dataset. 
We choose the axes such that the two points that are $2R$ apart lie along one of the axes. 
Thus, a $d$-dimensional hyper-sphere of radius $R$ would bound the dataset. 
In the worst case, e.g. with a distribution that fills the $d$-sphere, our axes will be such that $2R$ is the maximum pairwise distance along every axis. 
Such a distribution would also produce a balanced clustering.

Partition will select a maximally distant pair of points to use as poles, i.e., it will choose one of the $d$ axes along
which to split the cluster into two children. 
After one application of Cluster Partition, the maximum pairwise distance along that axis will be
bounded above by $R$. 
The next recursive Partition will select another of the $d$ axes. 
Thus, after at most $d$ applications of Cluster Partition, the
maximum pairwise distance along each axis will be bounded above by $R$. 
The overall (i.e., not restricted to be along one axis) maximum pairwise distance, 
will be bounded above by $R\sqrt{2}$ by, for example, two instances that lie at the extrema of different axes.

Thus, starting with a cluster $C$ of radius $R$, after at most $d$ Partitions, the descendants of $C$ will each have radius
bounded above by $\frac{r}{\sqrt{2}}$. In other words, cluster radii are guaranteed to decrease by a multiplicative factor of $\frac{\sqrt{2}}{2}$ after at 
most $d$ applications of Cluster Partition.

Note that, in practice, we never see a balanced clustering. Cluster Partition  produces unbalanced trees due to the varying density of the sampling 
in different regions of the manifold and the low dimensional "shape" of the manifold. Further, the cluster radii decrease by a factor much larger than 
$\frac{\sqrt{2}}{2}$ in practice, and often this decrease starts occurring after only one or two applications of Cluster Partition, rather than the 
upper bound of $d$ applications.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=2.5in]{images/geometry/sketchometry_final.png}
    \caption{Scaling Behavior of Radii in a Two-Dimensional Uniform-Density Disk}
    \label{fig:methods:scaling_behavior}
\end{figure}

\subsubsection {Complexity}
\label{subsubsec:methods:clustering:complexity}
%A lot of this is lifted from the CHESS paper and could probably get cut out. %

To analyze the computational complexity of hierarchical
clustering, we start by considering the top level of the hierarchy. 
For a data set with cardinality $n$, we have $n$ data points at this 
level, from which we randomly choose a sample of $\sqrt{n}$ points. 
We compute distances between every pair of seeds in this sample, 
thus making $n$ distance comparisons. Based on these distances 
we compute the center and left and right poles. Once the poles 
have been chosen, every data point except for the two poles is 
compared to both poles, for an additional 2$n$ comparisons.

In general, at depth $d$ of
the hierarchical tree, we have $\frac{n}{2^d}$ points per cluster (assuming a
perfectly balanced clustering), $\sqrt{\frac{n}{2^d}}$ seeds, $\frac{n}{2^d}$
pairwise seed comparisons, and $\frac{2n}{2^{d}}$ additional comparisons when 
each other data point is compared to the two poles, for a total of $\frac{3n}{2^d}$
distance comparisons. At each level, there are $2^d$ clusters,
so the total number of distance comparisons per level is
3$n$. Thus, the asymptotic complexity in terms of distance
comparisons for a cluster tree of depth $d$ is $\mathcal{O}(dn)$

By using an approximate partitioning with a $\sqrt{n}$ sample, we achieve $\mathcal{O}(n)$ cost of 
partitioning and $\mathcal{O}(n \log n)$ cost of building the tree. This is a significant improvement over 
exact partitioning and tree-building, which have $\mathcal{O}(n^2)$ and $\mathcal{O}(n^2 \log n)$ costs respectively.

\subsection {Index-Building and Sharded Search}
In this work, we also introduce the method of \emph{sharding} for $k$-nearest neighbors search.
With this approach, we aim to leverage the fact that $\rho$-nearest neighbors search is typically 
much faster than $k$-nearest neighbors search.

We begin by randomly partitioning the dataset into $s$ shards, where $s$ is determined by [TODO: how did 
we decide we're going to select $s$ for this version of the paper?]. On the first shard, we perform 
$k$-nearest neighbors search using one of the four algorithms described in Section [ADD SECTION: $k$-Nearest Neighbors Search].
The result of this search will give us the $k$ closest neighbors to the query in the first shard. We then take the furthest of 
those $k$ neighbors and use its distance from the query as the radius for $\rho$-nearest neighbors search (see section [ADD SECTION]) 
on each remaining shard. We then take the union of the results of these $\rho$-nearest neighbors searches, along with the $k$ neighbors
from the first shard, and then linear search over those points to find the actual $k$ nearest neighbors.

Note that the size of the shards is not necessarily consistent; it is often optimal to have the first shard be smaller than the others,
since the first shard is the only one for which we perform $k$-nearest neighbors search. However, making this first shard too small
creates the risk of significantly overestimating the radius, resulting in much slower $\rho$-nearest neighbors search on the remaining shards.
Finding the precise optimal ratio between the size of the first shard and the size of the remaining shards, as well 
as computing the expected level of error between the distance from the $k$-th closest neighbor in the first shard and the 
actual $k$-th closest neighbor, are topics for future work.

@@

In this section, we present four novel algorithms for $k$-nearest neighbors search: $k$-NN by Repeated $\rho$-NN, Sieve Search, and Sieve Search with Separate Centers, 
and  Greedy Sieve. 
We also have an implementation of linear search. We use a preliminary auto-tuning function to predict the variant which will perform 
best for a given query, dataset, and value of $k$. We then proceed with search using that variant.

In these algorithms, we typically use $H$, for \emph{hits}, to refer to the data structure which stores the closest points to the query found so far.

\subsubsection{$k$-NN by Repeated $\rho$-NN}
\label{subsubsec:methods:knn-search:repeated-rnn}
In this algorithm, we perform
$\rho$-nearest neighbors search at equal to the radius of the cluster tree divided by
the cardinality of the dataset, repeatedly increasing the search radius until $k$ neighbors
have been found.

Let $H$ be the set of nearest neighbors found thus far.
Search starts with a radius $m$ equal to the radius of the root cluster divided by
the cardinality of the dataset. We perform $\rho$-NN search with radius $m$. 
If no points are within a distance $m$ of the query, we increase the radius by a factor of 
2 and perform $\rho$-NN search again, repeating until at least one point is found, i.e., 
until $|H| > 0$.

Once $|H| > 0$, we continue to perform $\rho$-NN search, but instead of 
increasing the radius by a factor of 2 on each iteration, we increase it by a factor determined 
by the local fractal dimension of the clusters containing the points found so far. In particular, 
we increase the radius by a factor of 
\begin{equation} \text{argmin}\left(2, \left({\frac{k}{|H|}}\right)^{\frac{1}{\mu}}\right) \label{2} \end{equation}
where $\mu$ is the local fractal dimension (abbreviated lfd in algorithm below) in the region around the query point.
Intuitively, the factor by which we increase the radius should be \emph{inversely} related to the number of points found so far. 
Additionally, when the local fractal dimension at the radius scale from the previous iteration is low, this suggests that the data 
are relatively concentrated in that region; thus, the factor of radius increase should be \emph{directly} related to the 
local fractal dimension. Once $|H| >= k$, we return the $k$ closest points.

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated$\rho$-NN(\emph{tree, query, k})} % give the algorithm a caption
    \label{alg:knn-by-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[4] % enter the algorithmic environment
        \STATE $H \leftarrow$ $\emptyset$
        \STATE $m \leftarrow$ $\frac{root.radius}{root.cardinality}$
        \WHILE {$|H| = 0$}
            \STATE $H.push(\rho$-NN$(tree, query, m)$)
            \STATE $m \leftarrow 2m$
        \ENDWHILE
        \WHILE {$|H| < k$}
            \STATE $clusters \leftarrow \{ C: \exists p \in H \land p \in C \}$
            \STATE $\mu \leftarrow \frac{1}{|clusters|} \sum_{C \in clusters} C.lfd$
            \STATE $m \leftarrow \text{min}\left(2, \frac{k}{|H|^{\frac{1}{\mu}}}\right)$
            \STATE $H.push(\rho$-NN$(tree, query, m)$)
        \ENDWHILE
        \STATE $H.sort()$
        \STATE Return $H[0], H[1], \cdots H[k-1]$
    \end{algorithmic}
    \end{algorithm}

\subsubsection{Sieve Search}
\label{subsubsec:methods:knn-search:sieve}
With Sieve Search, we begin by letting $Q$ be a list initialized with the root cluster. This list will maintain the 
clusters and points which are still in contention for being one of the nearest neighbors.
While $Q$ contains at least one cluster (i.e., $Q$ is not a list of only points), we repeat the process described in 
the following paragraphs.

@@

We complete this process until $Q$ contains only points. At this point, we 
just use [INSERT ref to quicksort partition in paper] to find the $k$th 
nearest neighbor. Since this algorithm will also elements on the correct side of 
the index of the $k$th nearest neighbor, the first $k$ elements in $Q$ are the $k$-nearest neighbors.
